---
title: "Covid-19 by the Numbers"
author: "Josh Lewis"
date: "04/27/2023"
output: slidy_presentation
---

## Covid-19
  
  The Covid-19 pandemic was a devastating and unique event in our history. It created lasting transformations for society and still remains a controversial topic. Opinions range on masks, effectiveness of lock-down mandates, the lethality of the virus, and every facet of the pandemic.  
  This presentation will examine the data set of number of deaths of individuals infected with Covid-19 at the time of their death. This data set was produced by the U.S Department of Health & Human Services and is available at this  [link](https://catalog.data.gov/dataset/conditions-contributing-to-deaths-involving-coronavirus-disease-2019-covid-19-by-age-group). The data for 2020 and 2021 is final while 2022 and 2023 are provisional.  
  The presentation will contain another data set to supplement the Covid-19 data set with national and state wide population figures. The presentation will use the 2020 census from the United States Census Bureau for population figures, available at this [link](https://data.census.gov/table?q=populatoin+totals+by+state&tid=DECENNIALPL2020.P1).
  
#### Questions

* Which condition contributed to more deaths
* How age is related to deaths
* Which year had the most deaths
* Death-rate per month
* Total deaths per State
* Percentage of deaths per State

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries needed


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(usmap)
library(stringr)
library(knitr)
library(forcats)
```

### The data
```{r echo=FALSE}

# CSV provided by the U.S. Department of Health & Human Services available at data.gov
# Data set that contains the amount of people infected with COVID-19 at the time of their death and 
# underlying conditions that contributed to their death.
covid_d <- read.csv(
  "../data_sets/Conditions_Contributing_to_COVID-19_Deaths__by_State_and_Age__Provisional_2020-2023.csv",
  stringsAsFactors = FALSE
)
head(covid_d, 8)
```
***
### National total of deaths per pre-existing medical condition

```{r fig.width=7, fig.height=7, echo=FALSE}
# Data set shows each conditions grand total.
national <- covid_d %>% filter(
  State == "United States" & Age.Group == "All Ages"
  & Group == "By Total"
)
national$Condition.Group <- str_wrap(national$Condition.Group, width = 15)

national %>% group_by(Condition.Group) %>% summarise(total.deaths = sum(COVID.19.Deaths)) %>%
  ggplot(aes(Condition.Group, total.deaths)) +
  geom_col(fill = "#008080") + scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_classic() + labs(x = "Condition", y = "Total Deaths",
                         title = "Total Deaths per Condition") +
  theme(plot.title = element_text(hjust = 0.5))
```

***
#### Total Deaths by Condition and Age.

```{r echo=FALSE}
national_by_age <- covid_d %>% filter(State == "United States" & Group == "By Total" &
                                        Age.Group != "All Ages" & Age.Group != "Not stated")
national_by_age$Condition.Group <- str_wrap(national_by_age$Condition.Group, width = 15)
national_by_age %>% ggplot(aes(Condition.Group, COVID.19.Deaths, fill = Age.Group)) +
  geom_col() +  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_classic() + labs(x = "Condition", y = "Total Deaths",
                         title = "Total Deaths per Condition") +
  theme(plot.title = element_text(hjust = 0.5))

```

***
### National Total by Age.

```{r echo=FALSE}
national_by_age %>% group_by(Age.Group) %>% summarise(total.deaths = sum(COVID.19.Deaths)) %>% 
  ggplot(aes(Age.Group, total.deaths)) +
  geom_col(fill = "blue") + scale_y_continuous(labels = function(n){format(n, scientific = FALSE)})
```

***
### Fill by Condition Group.

```{r echo=FALSE}
national_by_age %>% ggplot(aes(Age.Group, COVID.19.Deaths, fill = Condition.Group)) +
  geom_col() + scale_y_continuous(labels = function(n){format(n, scientific = FALSE)})

```

***
### Monthly Totals and timeline Covid-19 deaths.

``` {r echo=FALSE, message=FALSE, warning=FALSE}
monthly_deaths <- covid_d %>% filter(State == "United States" & Age.Group == "All Ages" & Group == "By Month")
monthly_deaths_totals <- monthly_deaths %>% group_by(Start.Date) %>% summarise(total.month = sum(COVID.19.Deaths))
monthly_deaths_totals$Start.Date <- as.Date(monthly_deaths_totals$Start.Date, "%m/%d/%Y")
monthly_deaths_totals <- monthly_deaths_totals %>% arrange(Start.Date)
monthly_deaths_totals %>% ggplot(aes(Start.Date, total.month)) +
  geom_area(fill = "#800000") + theme_classic() + scale_y_continuous(labels = function(n){format(n, scientific = FALSE)})

monthly_age <- covid_d %>% filter(Group == "By Month" & State == "United States") %>% select(Start.Date, Condition, Age.Group, COVID.19.Deaths)
monthly_age$Start.Date <- as.Date(monthly_age$Start.Date, "%m/%d/%Y")
monthly_age %>% group_by(Start.Date, Age.Group) %>% summarise(Deaths = sum(COVID.19.Deaths)) %>%  filter(Age.Group != "All Ages") %>% 
  ggplot(aes(Start.Date, Deaths, fill = Age.Group)) +
  geom_area() + scale_y_continuous(labels = function(n){format(n, scientific = FALSE)})
```

***
### Yearly Stats

```{r echo=FALSE}
nation_by_year <- covid_d %>% filter(Group == "By Year" & State == "United States" & Age.Group != "All Ages"
                                     & Age.Group != "Not stated") %>% 
  select(Year, Condition.Group, Condition, Age.Group, COVID.19.Deaths)
nation_by_year$Condition.Group <- str_wrap(nation_by_year$Condition.Group, width = 15)
nation_by_year %>% ggplot(aes(Year, COVID.19.Deaths, fill = Condition.Group)) +
  geom_col()

```

***
### Timeline of total deaths.

```{r}
create_total_to_date <- function(data_V) {
  new_V <- vector("integer", length(data_V))
  new_V[1] = data_V[1]
  for(x in 2:length(data_V))
    new_V[x] = new_V[x - 1] + data_V[x]
  return(new_V)
}
```

```{r echo=FALSE}
running_total <- monthly_deaths_totals %>% mutate(
  to.date = create_total_to_date(total.month))
```

```{r echo=FALSE}
running_total %>% ggplot(aes(Start.Date, to.date)) +
  geom_line() + theme_classic() + theme_classic() + scale_y_continuous(labels = function(n){format(n, scientific = FALSE)}) + labs(x = "", y = "", title = "Cummulative Death Toll") +
  theme(plot.title = element_text(hjust = 0.5))
```

***
```{r echo=FALSE}
# Data set showing the total deaths per condition per state.
# Total is for all age-groups and dates 01/01/2020-03/18/2023.
states <- covid_d %>% filter(State != "United States" & Age.Group == "All Ages"
                             & Group == "By Total")

# Data set shows the total amount of deaths each state has
# incurred within dates 01/01/2020-03/18/2023.
state_total <- states %>% group_by(State) %>% 
  summarise(Total.Deaths = sum(COVID.19.Deaths)) %>% filter(State != "District of Columbia" & State != "Puerto Rico")
# Combine New York and New York City
state_total$Total.Deaths[state_total$State == "New York"] <-
  state_total$Total.Deaths[state_total$State == "New York"] +
  state_total$Total.Deaths[state_total$State == "New York City"]
state_total <- state_total %>% filter(State != "New York City")

kable(state_total %>% rename("Total Deaths" = Total.Deaths))


# Data set with the nations grand total of deaths.
national_total_deaths <- national %>% summarise(Died.With.Virus = sum(COVID.19.Deaths))
kable(national_total_deaths %>% rename("U.S. Total Deaths" = Died.With.Virus))
```

***
### The states

```{r fig.width= 8, fig.height= 8, echo=FALSE}
state_total %>%  ggplot(aes(Total.Deaths, reorder(State, Total.Deaths))) +
  geom_col(fill = "navy", width = .9) + scale_x_continuous(labels = function(n){format(n, scientific = FALSE)}) +
  labs(x = "", y = "", title = "Total Deaths per State") + theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

```

***
#### Top 9 States

```{r echo=FALSE, message=FALSE, warning=FALSE}
monthly_top_function <- function(dateV, numberV) {
  new_V <- vector("integer", length(dateV))
  for (x in 1:length(dateV)) {
    if(dateV[x] == "2020-01-01")
      new_V[x] = numberV[x]
    else
      new_V[x] = new_V[x - 1] + numberV[x]
  }
  return(new_V)
}

combine_yorks_monthly <- function(data_int_V, state_V) {
  new_V <- vector("integer", length(data_int_V))
  for (x in 1:length(data_int_V)) {
    if(state_V[x] == "New York")
      new_V[x] = data_int_V[x] + data_int_V[x + 1]
    else
      new_V[x] = data_int_V[x]
  }
  return(new_V)
}

# Need to add New York and New York City together.
top_10State_V <- vector("character", 10)
top_10State_V <- state_total %>% arrange(desc(Total.Deaths)) %>% slice(1:9) %>% .$State
top_10State_V[10] <- "New York City"
top_10 <- covid_d %>% filter(State %in% top_10State_V) %>%
  select(Start.Date, Group, State, Condition.Group, Condition, Age.Group, COVID.19.Deaths) %>% na.omit()
monthly_top_10 <- top_10 %>% filter(Group == "By Month" & Age.Group == "All Ages")
monthly_top_10 <- monthly_top_10 %>% group_by(Start.Date, State) %>% summarise(Total.Month = sum(COVID.19.Deaths))
monthly_top_10$Total.Month <- combine_yorks_monthly(monthly_top_10$Total.Month, monthly_top_10$State)
monthly_top_10$Start.Date <- as.Date(monthly_top_10$Start.Date, "%m/%d/%Y")
monthly_top_10 <- monthly_top_10 %>% arrange(State, Start.Date) %>% filter(State != "New York City")
t10V <- monthly_top_function(monthly_top_10$Start.Date, monthly_top_10$Total.Month)
monthly_top_10 <- monthly_top_10 %>% cbind(to.date = t10V)
```

```{r fig.height=7, fig.width=7, echo=FALSE}
monthly_top_10 %>% ggplot(aes(Start.Date, to.date, color = State)) +
  geom_line() + scale_y_continuous(labels = function(n){format(n, scientific = FALSE)}) +
  theme_classic()
```

***

```{r echo=FALSE}
# Function takes a column from a data set as a vector and vector of values that you wish to
# find. Returns a vector of the same length of data set containing true of false values.
# Which are the locations of the values being searched.
vectors_logical_OR <- function(data_V, condition_V) {
  t_f <- vector("logical", length(data_V))
  for (x in 1:length(data_V)) {
    for (v in 1:length(condition_V)) {
      if(data_V[x] == condition_V[v]){
        t_f[x] = TRUE
        break
      }
      if(v == length(condition_V)) {
        t_f[x] = FALSE
        break
      }
    }
  }
  return(t_f)
}

# CSV provided by the United States Census Bureau available at data.census.gov
# An official census is taken every ten years; this is the latest census from 2020.
census_data <- read.csv(
  "../data_sets/DECENNIALPL2020.P1-2023-04-13T175445.csv", stringsAsFactors = FALSE
)

# Data is in a wide format with each state being an attribute. state_census contains a row for each
# state and it's total population.
state_census <- census_data %>% filter(Label..Grouping. == "Total:") %>% pivot_longer(cols = !Label..Grouping., names_to = "States", values_to = "Total") %>% 
  select(-Label..Grouping.)
# Removes comma from numbers and converts column to a numeric value
state_census <- state_census %>% mutate(census = as.numeric(gsub(',', '', Total))) %>% 
  select(-Total)

# The following names contain a '.' instead of a face and will cause issues for future joins.
bad_names <- c("New.Hampshire", "New.Jersey", "New.Mexico", "New.York", "North.Carolina",
               "North.Dakota", "Rhode.Island", "South.Carolina", "South.Dakota",
               "West.Virginia")
# The Vector of the names needed for future joins.
correct_names <- c("New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina",
               "North Dakota", "Rhode Island", "South Carolina", "South Dakota",
               "West Virginia")

# Fixes the spelling of the state column.
state_vector <- state_census$States
t_f_vector <- vectors_logical_OR(state_vector, bad_names)
state_vector[t_f_vector] <- correct_names
state_census$States <- state_vector

# Data frame with ratio of population that died with covid in their system.
state_percentage <- left_join(state_total, state_census, join_by("State" == "States"))
state_percentage <- state_percentage %>% mutate(ratio = Total.Deaths / census)
percentage <- sprintf(state_percentage$ratio * 100, fmt = "%#.3f")
kable(cbind(state_percentage$State, percentage))
```

***
### State Percentages 

``` {r fig.width= 8, fig.height= 8, echo=FALSE}
state_percentage$ratio <- as.numeric(state_percentage$ratio)

state_percentage %>%  ggplot(aes(ratio, reorder(State, ratio))) +
  geom_col(fill = "navy", width = .9) + scale_x_continuous(labels = function(n){format(n, scientific = FALSE)}) +
  labs(x = "", y = "", title = "Percent of Deaths per State") + theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

```

***
#### US Map
```{r echo=FALSE}
us_total <- state_percentage %>% mutate(region = tolower(State))
us_total$ratio <- as.numeric(sprintf(us_total$ratio * 100, fmt = "%#.3f"))
map_states <- map_data("state")
map.df <- merge(map_states, us_total, by = "region", all.x=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long, y=lat, group=group)) +
  geom_polygon(aes(fill=Total.Deaths)) +
  geom_path() +
  scale_fill_continuous(name = "n", low = "lightblue", high = "darkblue", na.value = "grey90") +
  coord_map()
```

***
### US Map with Percentage

```{r echo=FALSE}
ggplot(map.df, aes(x=long, y=lat, group=group)) +
  geom_polygon(aes(fill=ratio)) +
  geom_path() +
  scale_fill_continuous(name = "%", low = "lightblue", high = "darkblue", na.value = "grey90") +
  coord_map()


```




